---
title: "Notes on 'Patterned and Plain Baked Clay in Pre-Pottery Contexts from Southeast Sulawesi, Indonesia' by Bulbeck et al."
author: "Ben Marwick"
date: "20 May 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE,
                      cache = TRUE)
```

# Introduction

In these notes I work through the paper to ensure I understand what is going on. I recompute some of the values, offer some observations where the ms. needs to be changed, and proposed some figures that we can add to the ms. 

# Data and Code

I would like to deposit our spreadsheets and my R code (i.e. something like this document) on a publicly accessible trustworty data repository. I have created a repository for this paper here: <http://doi.org/0.17605/OSF.IO/QF83X> and added mentions to it in the paper. There's nothing there yet, but I hope you will agree that this is worth to do. If we agree, I will upload the files at the same time that we submit to the journal. 

## Radiocarbon ages

I used the table of C14 ages in the ms. and calibrated them using the R pkg Bchron (IntCal13 curve).

```{r c14-read-in}
# read in data
library(here)

t_ages <- readxl::read_excel(here("data/data-aqy.2019.134/data/Talimbue_radiocarbon_ages.xlsx"))

# clean columns 
suppressPackageStartupMessages(library(tidyverse))
t_ages_clean <- 
t_ages %>% 
  mutate(Depth = if_else(str_detect(Depth, "cm"),
                         parse_number(Depth) / 100,
                         parse_number(Depth))) %>% 
  mutate(`Date (BP)` = str_remove_all(`Date (BP)`, ",")) %>% 
  separate(`Date (BP)`, into = c("age", "error")) %>% 
  mutate_at(vars(age, error), as.numeric)
```

Here is the calibration code:

```{r c14-calibrate, results='hide'}
# calibrate ages and plot
library(Bchron)

t_ages_clean_calibrated <- 
BchronCalibrate(t_ages_clean$age,
                t_ages_clean$error,
                calCurves = rep('intcal13', 
                                length(t_ages_clean$age)),
                positions = t_ages_clean$Depth)
```

I then fit a non-parametric chronology model to age/depth data according to the Compound Poisson-Gamma model defined by Haslett and Parnell (2008) <DOI:10.1111/j.1467-9876.2008.00623.x>

```{r c14-model, results='hide'}
t_out  = Bchronology(t_ages_clean$age,
                      t_ages_clean$error,
                      calCurves = rep('intcal13', 
                                      length(t_ages_clean$age)),
                      positions = t_ages_clean$Depth,
                     )
```

Here is the plot showing the interpolated ages:

```{r}
# plot with age-position 
plot(t_out,
     main="Talimbue Square B",
     xlab='Age (cal years BP)',
     ylab='Depth (m)',
     las=1, 
     dateHeight = 0.2,
     legLoc = "topright", 
     xlim = c(0, 20000))
```

I edited the plot by hand slightly to show the periods of rapid sedimentation, this is good because it clearly shows the uncertainties in our age-depth model

Using the chronology model computed above, I can predict ages for any depth in the site. We can use this to assign ages to the spits.  

```{r c14-predict, results='hide'}
# predict ages for a given depth
predictAges = predict(t_out, 
                      newPositions = seq(0,
                                         max(t_ages_clean$Depth),
                                         by=0.001))

predictAges_means <- colMeans(predictAges)

predictAges_means_df <- 
predictAges_means %>% 
  as_data_frame()  %>% 
  mutate(depth = as.numeric(str_remove(names(predictAges_means), "Pos"))) %>% 
  rename(age = value) 
```


## Spit depths and ages

I entered the spit depths from the recording sheets into an Excel file. I just used the depths recorded from the centre of the excavation square. For the lower depths that were not measured accurately I estimated a 0.05 m spit.  

```{r spit-depths}
# Square # depths for spits
t_sq_depths <- readxl::read_excel(here("data/data-aqy.2019.134/data/Spit depths from spit sheets.xlsx"))

knitr::kable(t_sq_depths)
```

# Densities of artefact and ecofact classes at Gua Talimbue Square B

We can plot this by depth or by age. Both are preferable to by Spit, because Spits are not 'real', for some peer-reviwers, at least.

```{r talimbue-sq-b-artefact-plot, fig.width=4, fig.show='hold'}
# Square B artefact densities (n/kg sediment)
t_arts_sq_B <- readxl::read_excel(here("data/data-aqy.2019.134/data/Talimbue_weights_analysis.xlsx"), sheet = "Talimbue_B_densities")

# Square # depths for spits
t_sq_depths <- readxl::read_excel(here("data/data-aqy.2019.134/data/Spit depths from spit sheets.xlsx"))

# join sq B
t_arts_sq_B_join <- 
  t_arts_sq_B %>% 
  left_join(t_sq_depths,  # get depths for spits
            by = c("Spit" = "spit")) %>% 
  # need 3 digits to match for the join
  mutate(sq_b_cumul_depth_below_surface = round(sq_b_cumul_depth_below_surface, 3)) %>% 
  left_join(predictAges_means_df,  # get ages for depths
            by = c("sq_b_cumul_depth_below_surface" = "depth"))

write_csv(t_arts_sq_B_join,
          here("data/data-aqy.2019.134/data/t_arts_sq_B_join.csv"))

# plot by depth 
t_arts_sq_B_join %>% 
  select(-Spit, 
         -SQB_center_end_level, 
         -SQE_center_end_level, 
         -age, 
         -sq_e_cumul_depth_below_surface) %>% 
  gather(material, density, -sq_b_cumul_depth_below_surface) %>% 
  ggplot(aes(sq_b_cumul_depth_below_surface, 
             density)) +
  geom_line() +
  ylab("Density (g/kg sediment)") +
  xlab("Depth (m)") +
  facet_wrap( ~ material, scales = "free_y", ncol = 1) +
  theme_minimal()

# plot by age 
t_arts_sq_B_join %>% 
  select(-Spit, 
         -SQB_center_end_level, 
         -SQE_center_end_level, 
         -sq_e_cumul_depth_below_surface,
         -sq_b_cumul_depth_below_surface) %>% 
  gather(material, density, -age) %>% 
  ggplot(aes(age/1000, 
             density)) +
  geom_line() +
  ylab("Density (g/kg sediment)") +
  xlab("Age cal BP (x 1000)") +
  facet_wrap( ~ material, scales = "free_y", ncol = 1) +
  theme_minimal()

```

# Densities of artefact and ecofact classes at Gua Talimbue Square E.

I think we should use the density-age plots for the paper. 

```{r talimbue-sq-e-artefact-plot, fig.width=4, fig.show='hold'}
# Square E artefact densities (n/kg sediment)
t_arts_sq_E <- readxl::read_excel(here("data/data-aqy.2019.134/data/Talimbue_weights_analysis.xlsx"), sheet = "Talimbue_E_densities")

# join sq E
t_arts_sq_E_join <- 
  t_arts_sq_E %>% 
  left_join(t_sq_depths,  # get depths for spits
            by = c("Spit" = "spit")) %>% 
  # need 3 digits to match for the join
  mutate(sq_e_cumul_depth_below_surface = round(sq_e_cumul_depth_below_surface, 3)) %>% 
  left_join(predictAges_means_df,  # get ages for depths
            by = c("sq_e_cumul_depth_below_surface" = "depth"))


# plot densities by depth
t_arts_sq_E_join %>% 
  select(-Spit, 
         -SQB_center_end_level, 
         -SQE_center_end_level, 
         -age, 
         -sq_b_cumul_depth_below_surface) %>% 
  gather(material, density, -sq_e_cumul_depth_below_surface) %>% 
  ggplot(aes(sq_e_cumul_depth_below_surface, 
             density)) +
  geom_line() +
  ylab("Density (g/kg sediment)") +
  xlab("Depth (m)") +
  facet_wrap( ~ material, scales = "free_y", ncol = 1) +
  theme_minimal()

# plot densities by age
t_arts_sq_E_join %>% 
  select(-Spit, 
         -SQB_center_end_level, 
         -SQE_center_end_level, 
         -sq_e_cumul_depth_below_surface,
         -sq_b_cumul_depth_below_surface) %>% 
  gather(material, density, -age) %>% 
ggplot(aes(age, 
           density)) +
  geom_line() +
  ylab("Density (g/kg sediment)") +
  xlab("Age cal BP (x 1000)") +
  facet_wrap( ~ material, scales = "free_y", ncol = 1) +
  theme_minimal()
```



# Correlation analysis of the Holocene artefact and ecofact densities (gm/kg of excavated sediment) from Gua Talimbue squares B and E

To check Table 2 in the ms. I explored some correlation and regression analyses. The plot below shows the Pearson's correlation coefficient values, they match what we see in the table. 

```{r}
Densities_for_regress_analysis <- readxl::read_excel("../data/Talimbue_weights_analysis.xlsx", 
                  sheet = "Densities_for_regress_analysis")

d <- Densities_for_regress_analysis %>% 
  select(Lithics, 
         Shellfish, 
         Vertebrate, 
         Plant, 
         Baked_clay, 
         Pottery)

# check the distributions of all pairs of variables. 
library(GGally)
ggpairs(d, progress = FALSE)
```

However, the plot above indicates that many of these variables are highly right-skewed. So we should use the more robust Spearman method for computing correlations, not Pearson:

```{r}
# should use spearman
library(corrr)
d %>% 
  correlate(method = "spearman") %>% 
  knitr::kable()
```

We can get p-values for the correlations, and we see that about half of them do not return significant correlations. This is contrary to the table in the ms. I think the problem is that we have different methods to compute the t-statistic. 

The t-statistic values I get here are quite different from in the ms. In the `Talimbue_weights_analysis.xlsx` spreadsheet, on sheet `Regression_analysis`, I see that you are computing the t-statistic using a formula like this `=D20*(139/(1-D20^2))` where `D20` is the cell with the *r* value, which we can write as

$$\frac {r(n-2)}{1-r^2}$$

. However, the [correct formula](https://onlinecourses.science.psu.edu/stat414/node/254/) for this is

$$\frac {r\sqrt{n-2}}{\sqrt{1-r^2}}$$

which in Excel, will look like `=(D20*SQRT(139))/SQRT(1-D20^2)`

What do you think? Do you have a source to confirm your method for computing the t-statistic? If it is wrote, we'll need to redo some tables and edit the text a bit more. 

Here is Pearson's correlation, with p-values:

```{r}
# cor.test with p-values 

var_pairs <- t(combn(names(d), 2)) %>%
  as_data_frame() %>% 
  setNames(c("x", "y"))

cor_test <- 
var_pairs %>% 
  dplyr::mutate(r.test = purrr::map2(x, y, 
                   ~ stats::cor.test(d[[.x]],
                                     d[[.y]], 
                                    method ="pearson")),
                r.test = purrr::map(r.test, broom::tidy)) %>%
  tidyr::unnest(r.test) %>% 
  select(-method, -alternative) %>% 
  mutate_if(is.numeric, round, 3) %>% 
  arrange(p.value)

knitr::kable(cor_test)
```

And here is Spearman, which I think is better for these data: 

```{r}
# cor.test with p-values 

var_pairs <- t(combn(names(d), 2)) %>%
  as_data_frame() %>% 
  setNames(c("x", "y"))

cor_test <- 
var_pairs %>% 
  dplyr::mutate(r.test = purrr::map2(x, y, 
                   ~ stats::cor.test(d[[.x]],
                                     d[[.y]], 
                                    method ="spearman")),
                r.test = purrr::map(r.test, broom::tidy)) %>%
  tidyr::unnest(r.test) %>% 
  select(-method, -alternative) %>% 
  mutate_if(is.numeric, round, 3) %>% 
  arrange(p.value)

knitr::kable(cor_test)
```

Let's also look at linear regressions for each pair of variables. We see only four of these combinations have significant relationships.

```{r}
# check linear models
list_of_models <- 
combn(names(d), 
      2, 
      function(x){lm(d[, x])}, simplify = FALSE)

vars <- combn(names(d), 2)
names(list_of_models) <- vars[1 , ]

table_of_lm_output <- 
map_dfr(list_of_models, broom::glance) %>% 
  mutate(var1 = vars[1, ]) %>% 
  mutate(var2 = vars[2, ]) %>% 
  select(var1, var2, everything()) %>% 
  mutate_if(is.numeric, round, 3)  %>% 
  arrange(p.value)

knitr::kable(table_of_lm_output)
```


# Vertical distribution of Gua Talimbue patterned baked-clay fragments.

I took the data in table 3 of the ms. and made this plot. It is very clear to see that curvilinear patterns are more typical of the I think this plot is much more effective at showing the trends that the table, and should replace the table. 

```{r fig.width=6}
t_vert_dist <- readxl::read_excel("../data/Talimbue_vertical_distribution_Gua_Talimbue_patterned_baked_clay_fragments.xlsx")

t_vert_dist %>% 
  mutate(Spits = str_glue('Spits {Spits}\n{age}')) %>% 
  select(Spits, ends_with("N")) %>% 
  gather(variable, value, -Spits) %>% 
  mutate(variable = str_remove_all(str_trim(variable),  "N")) %>% 
  ggplot(aes(Spits, 
             value)) +
  geom_col() +
  facet_wrap( ~ variable, ncol = 1, scales = "free_y") +
  theme_minimal() +
  ylab("Number of fragments") +
  xlab("Depositional phase")

```

