---
title: "Analying Talimbue Data"
output: 
  bookdown::html_document2:
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

Import and clean the data

```{r}
library(tidyverse)
library(kableExtra)

source("001-prepare-data.R")
source("002-clean-data.R")

db_data_retouch_subset <- 
  db_data_retouch %>% 
  filter(Material %in% c("Chert"))

bm_data_complete_flakes_all_tally <- 
  bm_data_complete_flakes_all %>% 
  group_by('Raw Material') %>% 
  tally() %>% 
  arrange(desc(n))

```       

Basic metric and technological properties of the assembalge

```{r}
# overview look at flakes to get a sense of variabitility of the entire
# assemblage 

# raw materials
tb_complete_flakes_clean %>% 
  group_by(Material) %>% 
  tally()

# metric dimensions only 
tb_complete_flakes_clean_metrics <- 
tb_complete_flakes_clean %>% 
  dplyr::select(
   `Oriented Length (mm)`,   
   `Oriented Width (mm)`,                       
   `Oriented Thick (mm)` ,        
   `Weight (Gram)` ,                                     
   `Striking Flat Form Width (mm)`,  
   `Striking Flat Form Thickness (mm)` ) %>% 
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "value") %>% 
  mutate(variable = case_when(
          variable == "Striking Flat Form Width (mm)" ~  "Striking Platform Width (mm)", 
          variable == "Striking Flat Form Thickness (mm)" ~  "Striking Platform Thickness (mm)",
          variable == "Oriented Thick (mm)" ~  "Oriented Thickness (mm)",
         TRUE ~ variable))
  
ggplot(tb_complete_flakes_clean_metrics) +
  aes(value) +
  geom_histogram() +
  facet_wrap( ~ variable, 
              scales = "free", nrow = 1) +
  xlab("") +
  ylab("") +
  theme_minimal(base_size = 14) +
  theme(panel.background = element_rect(fill='white',
                                        colour = "white"),
        plot.background = element_rect(fill='white',
                                        colour = "white"))

ggsave(here::here("figures/basic-flake-metrics-plot.png"), 
       h = 5, w = 20)

# and some of the techno attributes

tb_complete_flakes_clean_techno_nested_tallies <- 
tb_complete_flakes_clean %>% 
  dplyr::select(
   `Material`,   
   `Termination`,                       
   `Color` ,        
   `Cortex amount` ,                                     
   `Platform Preparation (Overhang)`,  
   `Heat treatment (potlids, crazing)` ) %>% 
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "value") %>% 
  filter(!is.na(value)) %>%
  filter(variable != "Color") %>% 
    mutate(value = case_when(
          value == "abent" ~  "absent", 
          value == "Absent" ~  "absent",
          value == "none" ~  "absent",
          value == "Andesit" ~  "Andesite",
         TRUE ~ value)) %>% 
      mutate(variable = case_when(
          variable == "Material" ~  "Raw material",
          variable == "none" ~  "absent",
          variable == "Andesit" ~  "Andesite",
         TRUE ~ variable)) %>% 
  nest(-variable) %>% 
  mutate(tallies = 
  map(data,  ~.x %>% 
        group_by(value) %>% 
        tally()
      )) 

tb_complete_flakes_clean_techno_nested_tallies_plot_list <- 
map2(tb_complete_flakes_clean_techno_nested_tallies$tallies,
     tb_complete_flakes_clean_techno_nested_tallies$variable,
    ~ggplot(.x) +
      aes(value, n) +
      geom_col() +
      xlab("") +
      ylab("") +
      theme_minimal(base_size = 14) +
      theme(panel.background = element_rect(fill='white',
                                        colour = "white"),
        plot.background = element_rect(fill='white',
                                        colour = "white")) +
      ggtitle(.y) +
      scale_x_discrete(guide = guide_axis(n.dodge = 2))
      )      

library(cowplot)
plot_grid(plotlist = tb_complete_flakes_clean_techno_nested_tallies_plot_list,
          nrow = 1
         )

ggsave(here::here("figures/basic-flake-techno-plot.png"), 
       h = 5, w = 20)
      
```

Investigate differences between depositional phases


```{r}

# dep phase by flake mass by depositional unit
dep_phase_mass_plot <- 
ggplot(tb_complete_flakes_clean) +
  aes(depositional_phase, 
      `Weight (Gram)`) +
  geom_boxplot()+
  geom_quasirandom(alpha = .2) +
  scale_y_log10(labels = scales::label_comma(accuracy = 0.1)) +
  scale_x_discrete(labels = c('Phase 1', 
                              'Phase 2',
                              'Phase 3',
                              'no phase')) +
  xlab("") +
  ylab("Mass (g)") +
  theme_minimal(base_size = 14) +
  theme(panel.background = element_rect(fill='white',
                                        colour = "white"),
        plot.background = element_rect(fill='white',
                                        colour = "white"))

aov( `Weight (Gram)`  ~ depositional_phase, 
     data =  tb_complete_flakes_clean %>% 
       filter(depositional_phase != "no_phase")) %>% 
  summary()


# dep phase by termination

dep_phase_termination_plot <- 
ggplot(tb_complete_flakes_clean %>% 
         filter(!is.na(Termination))) +
  aes(depositional_phase, 
     fill =  Termination) +
  geom_bar(position = "fill")  +
  scale_fill_viridis_d() +
  scale_x_discrete(labels = c('Phase 1', 
                              'Phase 2',
                              'Phase 3',
                              'no phase')) +
  xlab("") +
  ylab("") +
  theme_minimal(base_size = 14) +
  theme(panel.background = element_rect(fill='white',
                                        colour = "white"),
        plot.background = element_rect(fill='white',
                                        colour = "white"))


# dep phase by cortex
dep_phase_cortex_plot <- 
ggplot(tb_complete_flakes_clean %>% 
         filter(!is.na(`Cortex amount`))) +
  aes(depositional_phase, 
     fill =  `Cortex amount`) +
  geom_bar(position = "fill")   +
  scale_fill_viridis_d() +
  scale_x_discrete(labels = c('Phase 1', 
                              'Phase 2',
                              'Phase 3',
                              'no phase')) +
  xlab("") +
  ylab("") +
  theme_minimal(base_size = 14) +
  theme(panel.background = element_rect(fill='white',
                                        colour = "white"),
        plot.background = element_rect(fill='white',
                                        colour = "white"))

plot_grid(dep_phase_mass_plot,
          dep_phase_termination_plot,
          dep_phase_cortex_plot,
          nrow = 1
          )

ggsave(here::here("figures/phases-flake-panel.png"), 
       h = 5, w = 20)
```

Explore patterns in the assemblage using PCA

Looking at the correlations between the variables in BM's Data Counts

```{r}
library(FactoMineR)

ggpairs(bm_data_counts,
        lower = list(continuous = wrap("smooth")))
```

The Scatterplot matrix shows several correlated variables and some clusters. Will attempt a PCA to better understand the structure of this data set.

```{r}
res_pca <- PCA(bm_data_counts %>% 
                 select(-Spit, 
                        -Total,
                        -`Non-artefact`,
                        -`Unidentified/questions`),
               graph = FALSE)

library(factoextra)
fviz_screeplot(res_pca)
```

Here's the biplot

```{r}
fviz_pca_biplot(res_pca,
                repel = TRUE,
                title = "PCA Biplot for Phases 2 and 3") +
  theme_minimal(base_size = 14) +
  theme(panel.background = element_rect(fill='white',
                                        colour = "white"),
        plot.background = element_rect(fill='white',
                                        colour = "white"))

ggsave(here::here("figures/phases-counts-pca.png"), 
       h = 5, w = 5)
```

Figure 1: Visualzation of Eigenvalues

Figure 1 is a visulatizaton of our Eigenvalues. We can see that dimensions 1 and 2 are contributing the most to the variability.

```{r}
fviz_contrib(res_pca,
             choice = "var",
             axes = 1,
             top = 10)

fviz_contrib(res_pca,
             choice = "var",
             axes = 2,
             top = 10)
```

Broken Flakes, Complete Flakes, and Cores are contributing the most to the variability of this data set. Retouched pieces makes up a minimal amount of variability.

Figure 3: Contributions to variables to each dimension

Figure 3 shows each contribution of our variables that go into each dimension. Retouched pieces have a positive contributions for Dimensions 2 and 3.


```{r}
res_pca$var$coord %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "variable") %>% 
  as_tibble() %>% 
  pivot_longer(-variable,
               names_to = "Dims",
               values_to = "values") %>% 
  filter(Dims %in% c("Dim.1", "Dim.2", "Dim.3")) %>% 
  ggplot()+
  aes(values,
      variable,
      fill = variable)+
  geom_col(show.legend = FALSE) +
  facet_wrap( ~ Dims) +
  labs(y=NULL) +
  theme_minimal(base_size = 16) +
  theme(panel.background = element_rect(fill='white',
                                        colour = "white"),
        plot.background = element_rect(fill='white',
                                        colour = "white"))

ggsave(here::here("figures/phases-counts-pca-contrib.png"), 
       h = 7, w = 5)
```




```{r}

# pca_grouping by flake mass by depositional unit
ggplot(tb_complete_flakes_clean) +
  aes(pca_group, 
      `Weight (Gram)`) +
  geom_boxplot()+
  geom_quasirandom(alpha = .85) +
  scale_y_log10()

aov( `Weight (Gram)` ~ pca_group, 
     data =  tb_complete_flakes_clean %>% 
       filter(depositional_phase != "no_phase")) %>% 
  summary()

# pca_grouping by  Oriented Thick (mm)
ggplot(tb_complete_flakes_clean) +
  aes(pca_group, 
      `Oriented Thick (mm)`) +
  geom_boxplot()+
  geom_quasirandom(alpha = .85) +
  scale_y_log10()

aov( `Oriented Thick (mm)`  ~ pca_group, 
     data =  tb_complete_flakes_clean %>% 
       filter(depositional_phase != "no_phase")) %>% 
  summary()

# pca by termination
ggplot(tb_complete_flakes_clean) +
  aes(pca_group, 
     fill =  Termination) +
  geom_bar(position = "fill")  

```




```{r}
# looking into retouch
library(GGally)
library(ggbeeswarm)

hist(as.numeric(bm_data_retouch$`Platform Width`))

hist(bm_data_counts$`Retouched piece`)

hist(db_data_retouch$`Maksimal Length( mm)`)

ggplot(db_data_retouch_subset) +
  aes(Spit, 
      `Maksimal Length( mm)`) +
  geom_boxplot()+
  geom_quasirandom(alpha = .85)

# change in frequency of retouch over time

# by spit
ggplot(bm_data_counts) +
  aes(Spit, 
      `Retouched piece`
      ) +
  geom_col() +
  theme_minimal()

# by depositional phase
tb_retouch %>% 
  filter(Material == "Chert") %>% 
  group_by(depositional_phase) %>% 
  tally() %>% 
  ggplot() +
  aes(depositional_phase, n) +
  geom_col() +
  theme_minimal()
```
